Think deep for reef resilience

### 2025-08-08 paper/paper_draft.qmd
Expand DHW description
Add reference to table.

@eco-ben to address comments :)

Revert change to table

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled 1km
resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4 and are a part of the Sixth Phase of the
Coupled Model Intercomparison Project (CMIP6) based on evaluation conducted by Grose et al.
[-@grose2023]. These GCMs were found to rank highly in the recreation of historical
atmospheric temperature and precipitation conditions in the Australasian region. While this
assessment does not consider oceanic data, Grose et al. [-@grose2023] provide a
region-specific selection mechanism which we use as an indicator of general performance of
GCMs for the region.

Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology
[@liu2017; @skirving2020]. For each pixel of SST data, the monthly mean SST climatology (MM)
was calculated as the mean SST for each month over 1985 to 2012, adjusted to centre at year
1988.2857 (see Skirving et al. [-@skirving2020]). The "HotSpot SST Anomaly" [@skirving2020]
was calculated as the difference between daily SST values and the maxmimum of a pixel's MM
climatology (maximum monthly mean). DHW values were then calculated as the 12-week rolling
sum of positive HotSpot SST Anomaly values for a pixel. Within each year the maximum of
monthly mean DHW values was calculated. Reef scale DHW timeseries were extracted by taking
the average DHW level over pixels touching a reef.

The five selected GCMs span a range of Equilibrium Climate Sensitivity (ECS) values within
the Intergovernmental Panel on Climate Change (IPCC) assessed likely range, and GCM
development families. The IPCC defines ECS as the long term warming caused by a doubling of
carbon dioxide above its pre-industrial concentration [@forster2021]. The five GCMs
therefore allow for a range of possible climate futures to be explored, capturing the
uncertainty in future responses to CO2 increases (@sfig-ecs; @fig-methods-dhws; [@grose2023;
@shepherd2018]).

Shared Socio-economic Pathway 2-4.5 (approximate to the previously used RCP 4.5) was used to
represent the plausible emission pathway under current emission trajectories, with similar
atmospheric CO2 levels to present day remaining until 2050 [@hausfather2020; IPCC
-@climate2023]. This SSP trajectory estimates 2.7°C of global atmospheric warming in the
2081-2100 period relative to 1850-1900.

\[**TI note: The above needs to be expanded a little. IPCC has not been mentioned before at
this point either**\]

\[**TI: Needs a table clearly communicating each GCM, ECS value, and what each represent
(e.g., usual, optimistic scenario, etc)**\]

| Global Climate Model source | ECS value (°C) | Interpreted relative severity | Individual results displayed |
|-----------------------|-----------------------|-----------------------|-----------------------|
| EC-Earth3-Veg | 4.31 | High | **insert repo citation** |
| ACCESS-CM2 | 4.72 | High | **insert repo citation** |
| ACCESS-ESM1-5 | 3.87 | Intermediate | Main text |
| GFDL-CM4 | 2.9 | Low | **insert repo citation** |
| NorESM2-MM | 2.5 | Low | **insert repo citation** |

: Table 1: Information for each GCM used in this investigation including ECS value,
interpreted relative severity and where results are individual GCM results are displayed.
GCMs in context with assessed plausible ranges of ECS values are displayed in [@fig-ecs].
ACCESS-ESM1-5 is taken as an intermediate model based on the range of ECS values, when
assessing factors impacting reef resilience these results are indicative of results from all
GCMs assessed. Therefore, inidividual results for other GCMs are available at **insert repo
citation**. ECS value sources: .

@@

Under each GCM, in all management areas, low, medium and high cluster reefs have low
differences between DHW trajectories (@sfig-gcm-dhw-trajectories) \[**TI:** **I don't
understand this sentence - what is meant by reefs have low differences?**\]. The highest
peak DHW levels occur under GFDL-CM4 between 2045 and 2050, while greater increases in
overall DHW timeseries occur under ACCESS-CM2.

@@

Our findings on the importance of connectivity may differ due to the highly uncertain nature
of larval connectivity modelling [@vogt-vincent2023], the way larval flows and juvenile
recruitment are implemented in ADRIA - CoralBlox, and the magnitude of early bleaching
events decreasing the breeding capacity of strongly connected reefs. Analysing daily
connectivity data Vogt-Vincent et al. [-@vogt-vincent2023] suggest that time-mean
connectivity may not be suitable for capturing the seasonal - ecological time scale
connectivity between locations due to high daily variability in connectivity data. This is
supported by the connectivity data used in our investigation, which have large relative
standard deviations for connectivity between reefs over the 24 days used
(@sfig-connectivity-rsd). Using alternative summarisation methods that capture
ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Further improvements and research in connectivity modelling are
necessary to provide a robust comparison of all factors influencing reef resilience.

>>>>

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled 1km
resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GFDL-CM4 and are a part of the Sixth Phase of the
Coupled Model Intercomparison Project (CMIP6) based on evaluation conducted by Grose et al.
[-@grose2023]. These GCMs were found to rank highly in the recreation of historical
atmospheric temperature and precipitation conditions in the Australasian region. While this
assessment does not consider oceanic data, Grose et al. [-@grose2023] provide a
region-specific selection mechanism which we use as an indicator of general performance of
GCMs for the region. An overview of the selected GCMs are provided in @tbl-gcms. The process
described herein was conducted for all reefs within the GBR Marine Park to create DHW
projections spanning from 2025 to 2075.

The five selected GCMs span a range of Equilibrium Climate Sensitivity (ECS) values within
the Intergovernmental Panel on Climate Change (IPCC) assessed likely range, and GCM
development families \[**TI: Ben, what does "GCM development families" mean?**\]. The IPCC
defines ECS as the long term warming caused by a doubling of carbon dioxide above its
pre-industrial concentration [@forster2021]. The five GCMs therefore allow for a range of
possible climate futures to be explored, capturing the uncertainty in future responses to
CO2 increases (@fig-ecs; @sfig-methods-dhws; [@grose2023; @shepherd2018]) **\[TI: Missing
figure reference?\]**. Projections using Shared Socio-economic Pathway (SSP) 2-4.5 was used
to represent the plausible emission pathway. It represents a moderate emissions scenario
with continued CO2 emissions comparable to present day until 2050, after which emissions
gradually decline. Under this scenario, atmospheric concentrations approximately reach
540ppm by 2100 [@hausfather2020; IPCC -@climate2023]. The selected GCMs represent 2.5°C to
4.31°C of global atmospheric warming in the 2081-2100 period relative to 1850-1900.

DHW climatological baselines were calculated from the SST data following NOAA methodology
[@liu2017; @skirving2020]. For each pixel of SST data, the Monthly Mean SST climatology was
calculated by computing the mean SST value for each calendar month (January through
December) across the 28-year baseline period from 1985 to 2012. This climatological baseline
was then temporally adjusted to center at year 1988.2857, which represents the midpoint of
the reference period and accounts for any long-term trends within the baseline years
themselves. The adjustment ensures that the climatology represents conditions at a specific
temporal reference point rather than being influenced by gradual warming or cooling trends
that may have occurred during the 1985-2012 period. The resulting monthly climatologies
serve as a standardized reference against which current SST conditions can be compared to
identify thermal anomalies and assess coral bleaching risk, following the methodology
established by Skirving et al. [-@skirving2020].

**\[TI: Ben, could you provide more details in the paragraph above please? What resolution
is the data in? 10m\^2?\]**

Degree Heating Weeks (DHWs) were then derived using the "HotSpot SST Anomaly" approach
[@skirving2020]. The "HotSpot SST Anomaly" represents the thermal stress experienced by
coral reefs and is calculated as the difference between daily SST values and the maximum of
a pixel's monthly mean climatology across all twelve months (i.e., the maximum monthly mean
temperature). This approach uses the annual maximum temperature as the baseline because
corals are adapted to withstand their local warmest conditions under normal seasonal cycles.
When temperatures exceed this seasonal maximum, corals experience thermal stress that can
lead to bleaching. Only positive anomalies are considered relevant for coral stress, as
temperatures below the seasonal maximum do not contribute to bleaching risk.

DHW values on a per-pixel basis were then calculated as the 12-week rolling sum of positive
HotSpot SST Anomaly values, expressed in units of °C-weeks. The 12-week accumulation period
reflects the cumulative nature of coral thermal stress. Brief temperature spikes may be
tolerable, but sustained or repeated exposure to elevated temperatures causes progressively
greater physiological stress and bleaching severity. DHW values integrate both the magnitude
and duration of thermal stress: a moderate anomaly (e.g., 1°C) sustained for 12 weeks
produces the same DHW value as a severe anomaly (e.g., 3°C) lasting 4 weeks, though the
biological impacts may differ. DHW estimates for each indicative reef (see @greatba2022)
were then determined by taking the mean DHW value for all pixels within each reef outline.

**\[TI: Ben, I did my best to expand on DHW calculations and detail HotSpot SST anomaly,
could you please check?\]**

::: {#tbl-gcms}
| Global Climate Model source | ECS value (°C) | Interpreted relative severity | Results found in |
|-----------------------------|----------------|-------------------------------|------------------|
| EC-Earth3-Veg               | 4.31           | High                          | **insert repo citation**    |
| ACCESS-CM2                  | 4.72           | High                          | **insert repo citation**    |
| ACCESS-ESM1-5               | 3.87           | Intermediate                  | Main text        |
| GFDL-CM4                    | 2.9            | Low                           | **insert repo citation**    |
| NorESM2-MM                  | 2.5            | Low                           | **insert repo citation**    |

: Information for each GCM used in this investigation including ECS value, interpreted
relative severity and where results are individual GCM results are found. GCMs in context
with assessed plausible ranges of ECS values are displayed in [@sfig-ecs] **\[TI: Missing
figure?\]**. ACCESS-ESM1-5 is taken as an intermediate model based on the range of ECS
values, thus results are displayed in the main text. ECS value sources: .
:::

@@

Under each GCM, in all management areas, low, medium and high cluster reefs have low
differences between DHW trajectories (@fig-gcm-dhw-trajectories) \[**TI:** **I don't
understand this sentence - what is meant by reefs have low differences?**\]. The highest
peak DHW levels occur under GFDL-CM4 between 2045 and 2050, while greater increases in
overall DHW timeseries occur under ACCESS-CM2.

![Reef Degree Heating Week timeseries clustered within each management area, for each GCM
analysed. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover
between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with
intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’
clusters (blue) are clusters with the highest median area-relative cover between 2030 and
2040 within each bioregion. Solid lines represent the cluster median timeseries and bands
represent the 2.5 to 97.5 percentile confidence intervals. Each management area is
represented five times (for each GCM), with each GCM displayed
horizontally.](../outputs/ADRIA_results/HighResCoralStress/figs/GCM_dhw_timeseries_plot.png){#fig-gcm-dhw-trajectories
width="14.82cm"}

@@

Our findings on the importance of connectivity may differ due to the highly uncertain nature
of larval connectivity modelling [@vogt-vincent2023], the way larval flows and juvenile
recruitment are implemented in ADRIA - CoralBlox, and the magnitude of early bleaching
events decreasing the breeding capacity of strongly connected reefs. Analysing daily
connectivity data Vogt-Vincent et al. [-@vogt-vincent2023] suggest that time-mean
connectivity may not be suitable for capturing the seasonal - ecological time scale
connectivity between locations due to high daily variability in connectivity data. This is
supported by the connectivity data used in our investigation, which have large relative
standard deviations for connectivity between reefs over the 24 days used
[@sfig-connectivity-rsd]. Using alternative summarisation methods that capture
ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Further improvements and research in connectivity modelling are
necessary to provide a robust comparison of all factors influencing reef resilience.
